---
title: "Getting Started with fine-mapping-benchmark"
author: "Lucas Siu"
date: "`r format(Sys.Date(), '%B %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    df_print: paged
vignette: >
  %\VignetteIndexEntry{Getting Started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
# Run every chunk from the project root, not from vignettes/. This means
# (a) `source("R/...")` paths work without ../ prefix, (b) the renv
# autoloader activates because it walks up from the cwd looking for
# renv/, and (c) `data/`-relative paths inside simulator helpers resolve
# correctly.
knitr::opts_knit$set(root.dir = "..")

knitr::opts_chunk$set(
  echo       = TRUE,
  message    = FALSE,
  warning    = FALSE,
  cache      = TRUE,
  fig.align  = "center",
  fig.width  = 6,
  fig.height = 4
)
```

# 1. Introduction

**Fine-mapping** is the task of identifying which variant(s) within a
GWAS-significant locus are *causally* associated with a trait, as opposed to
merely correlated with the causal variant through linkage disequilibrium (LD).
Several methods have been developed for this — they take summary statistics
(z-scores or β̂ / SE) plus an LD matrix and return a *posterior inclusion
probability* (PIP) and one or more *credible sets* of candidate causal
variants.

This package provides a unified framework for **benchmarking** those methods
on simulated data with known ground truth. It supports:

- **Simulation** of genotypes from real 1000 Genomes haplotypes and phenotypes
  under sparse or sparse+infinitesimal genetic architectures, optionally with
  functional annotations.
- **A catalogue of methods** with a standardised input/output interface —
  SuSiE, SuSiE-inf, FINEMAP, CARMA, ABF, PAINTOR, Funmap, BEATRICE, Functional
  BEATRICE, plus baselines (`marginal_z`) and PolyFun-style comparators
  (`polyfun_oracle`, `polyfun_est`).
- **Evaluation** via AUPRC, credible-set coverage/power/size, PIP calibration,
  with stratification by S (causal count), φ (heritability), p_causal,
  causal-variant MAF, and the simulation's annotation regime.
- **Plotting** of stratified results to a multi-page PDF.

This vignette walks you through a small end-to-end run that uses only Tier 1
(pure R) methods and the sim1000G example VCF bundled with the
`sim1000G` package — so you don't need to download anything beyond R packages.

If you want to use the external methods (FINEMAP, PAINTOR, Funmap, BEATRICE),
see the
[`README` installation tiers](../README.md#installation).

---

# 2. Setup

Source the project files. The Tier 1 set is enough for SuSiE, ABF, CARMA, the
baselines, and PolyFun-style methods.

```{r sources}
source("R/utils.R")
source("R/simulate_genotypes.R")
source("R/simulate_phenotypes.R")
source("R/run_simulation.R")
source("R/run_methods.R")
source("R/evaluate.R")
source("R/plot_results.R")
source("R/wrapper_susie.R")
source("R/wrapper_abf.R")
source("R/wrapper_marginal_z.R")
```

> **Note on paths.** All chunks in this vignette run from the **project
> root** — set in the setup chunk via
> `knitr::opts_knit$set(root.dir = "..")`. The renv autoloader needs this
> to find the project's R library. If you copy the chunks into a script,
> run R from the project root and the paths above work as-is.

The Tier 1 dependencies (`susieR`, `sim1000G`, `ggplot2`, `gridExtra`,
`CARMA`) install via `renv::restore()` from the project root. See
[README §Installation](../README.md#installation) if you have not done this
yet.

---

# 3. Simulate a small dataset

`run_simulation()` orchestrates everything: it simulates genotypes from
1000 Genomes haplotypes (just once), then iterates over a grid of
parameter combinations to produce one phenotype + summary-statistic set per
combination.

We use small parameters so this knits in under a minute:

```{r simulate}
sim <- run_simulation(
  n_regions = 2,        # 2 independent genomic regions
  n         = 200,      # 200 simulated individuals
  p         = 80,       # ~80 SNPs per region
  n_iter    = 2,        # 2 phenotype replicates per parameter combination
  S         = c(1, 2),  # sweep over 1 and 2 causal variants per region
  phi       = c(0.2, 0.4),  # sweep over 20% and 40% PVE
  model     = "sparse",
  annotations = "none",
  vcf_files = NULL,     # use the bundled sim1000G example VCF
  seed      = 42,
  verbose   = FALSE
)
```

Total scenarios: `length(S) × length(phi) × n_iter = 2 × 2 × 2 = 8`,
each containing `n_regions = 2` per-region datasets — so 16 fits per method.

---

# 4. Inspect the simulation object

The result has three top-level pieces:

```{r sim-toplevel}
names(sim)
```

### `sim$genotypes`

A list of length `n_regions`, one per region. Each element holds the genotype
matrix, MAFs, LD matrix, etc.

```{r sim-genotypes}
length(sim$genotypes)
str(sim$genotypes[[1]], max.level = 1)
```

The LD matrix is the same across all scenarios for a given region (genotypes
are simulated once and reused):

```{r sim-ld}
dim(sim$genotypes[[1]]$LD)
sim$genotypes[[1]]$LD[1:4, 1:4]
```

### `sim$scenarios`

A list of length `n_scenarios`, one per (S, φ, iter) combination. Each
scenario has the per-region phenotype + summary statistics + truth.

```{r sim-scenarios}
length(sim$scenarios)
str(sim$scenarios[[1]], max.level = 1)
```

The ground truth for one (scenario × region) sits in `truth`:

```{r sim-truth}
truth <- sim$scenarios[[1]]$regions[[1]]$truth
truth$causal_indices    # which variants are causal
truth$causal_effects    # their true effect sizes
truth$pve               # realised proportion of variance explained
```

### `sim$params`

All the parameters used, recorded for reproducibility — useful when loading a
saved `.rds` later:

```{r sim-params}
str(sim$params, max.level = 1)
```

---

# 5. Run fine-mapping methods

`run_methods()` applies one or more methods to every (scenario × region)
combination. We run three Tier 1 methods here: SuSiE, ABF, and `marginal_z`
(the model-free baseline).

```{r run-methods}
out <- run_methods(
  simulation  = sim,
  methods     = c("susie", "abf", "marginal_z"),
  method_args = list(
    susie      = list(L = 5, coverage = 0.95),
    abf        = list(prior_variance = 0.04),
    marginal_z = list(coverage = 0.95)
  ),
  save        = FALSE,
  verbose     = FALSE
)
```

The structure mirrors the methods you asked for:

```{r out-toplevel}
names(out)
out$methods_run
```

Each method's results live under its key:

```{r out-method-summary}
cat(sprintf("SuSiE     : %d/%d fits OK, %.2fs total\n",
            out$susie$n_total - out$susie$n_failed,
            out$susie$n_total, out$susie$total_runtime_seconds))
cat(sprintf("ABF       : %d/%d fits OK, %.2fs total\n",
            out$abf$n_total - out$abf$n_failed,
            out$abf$n_total, out$abf$total_runtime_seconds))
cat(sprintf("marginal_z: %d/%d fits OK, %.2fs total\n",
            out$marginal_z$n_total - out$marginal_z$n_failed,
            out$marginal_z$n_total, out$marginal_z$total_runtime_seconds))
```

---

# 6. Inspect a fit

Each per-fit result follows a standard format. Pulling one SuSiE fit:

```{r out-one-fit}
fit <- out$susie$results[[1]]
str(fit, max.level = 1)
```

### The PIP vector

`fit$pip` is a numeric vector of length `p` (the number of SNPs in that
region) — posterior inclusion probability for each variant.

```{r out-pip}
length(fit$pip)
range(fit$pip)
head(sort(fit$pip, decreasing = TRUE), 5)   # top-5 variants
```

### Credible sets

`fit$credible_sets` is a list of integer vectors. Each vector contains the
1-based indices of variants in one credible set. SuSiE can return multiple
credible sets (one per "single-effect component"); ABF and `marginal_z` return
just one per region.

```{r out-cs}
fit$credible_sets
```

Compare against the truth (causal indices we recorded earlier):

```{r out-cs-vs-truth}
truth$causal_indices   # the true causal variant index/indices
fit$credible_sets[[1]] %in% truth$causal_indices
```

### Method-specific outputs

`fit$additional` holds method-specific extras. For SuSiE this includes the
per-component posterior matrix, convergence diagnostics, and the ELBO trace:

```{r out-additional}
names(fit$additional)
fit$additional$converged
fit$additional$n_iter_run
```

---

# 7. Evaluate

`evaluate_methods()` computes AUPRC, credible-set coverage/power/size, PIP
calibration, and runtime — pooled globally and stratified along several
axes.

```{r evaluate}
eval <- evaluate_methods(
  simulation = sim,
  results    = out,
  save       = FALSE,
  verbose    = FALSE
)
```

### Global metrics

Each method has a `$global` entry with the pooled-across-everything numbers:

```{r eval-global}
str(eval$susie$global, max.level = 1)
```

A compact one-row-per-method summary:

```{r eval-summary-table}
fmt <- function(x, d = 3) ifelse(is.na(x), "NA", formatC(x, format = "f", digits = d))
do.call(rbind, lapply(eval$methods_evaluated, function(m) {
  g <- eval[[m]]$global
  data.frame(
    method        = m,
    auprc         = fmt(g$auprc),
    cs_coverage   = fmt(g$cs_coverage),
    cs_power      = fmt(g$cs_power),
    cs_size_med   = fmt(g$cs_size_median, 1),
    runtime_s     = fmt(g$runtime_mean, 2),
    stringsAsFactors = FALSE
  )
}))
```

You should see SuSiE and ABF appreciably above `marginal_z` — that's the
expected ordering, since `marginal_z` does no modelling at all.

### Stratified metrics

Several axes are computed automatically. By number of causal variants:

```{r eval-by-S}
names(eval$susie$by_S)
cat("\nAUPRC by S, SuSiE:\n")
sapply(eval$susie$by_S, function(x) round(x$auprc, 3))
```

By minimum causal MAF (added in Phase 2):

```{r eval-by-maf}
names(eval$susie$by_causal_maf)
cat("\nAUPRC by causal MAF bin, SuSiE:\n")
sapply(eval$susie$by_causal_maf, function(x) round(x$auprc, 3))
```

> **Why only one bin here?** The bundled `sim1000G` VCF contains only common
> variants (MAF > 0.05), so all causals fall in the `"common"` bin. With the
> real 1000G windows downloaded by `inst/scripts/prepare_vcfs.R` (or your own
> VCFs), the `"rare"` and `"low"` bins start to appear.

Other axes available: `by_phi`, `by_p_causal` (sparse_inf simulations only),
`by_true_annotation_type` (when comparing across annotation regimes).

---

# 8. Plot

`plot_results()` writes a multi-page PDF with PR curves, PIP calibration
plots, summary tables, and stratified-metric panels.

```{r plot, eval = FALSE}
plot_results(eval, output_file = "getting-started-results.pdf",
             save = TRUE, verbose = FALSE)
```

The PDF has, in order:

1. **Global PR curves** for all methods on one panel
2. **Global PIP calibration** faceted by method
3. **Global summary table** with means ± SE across replicates
4. **By-S grids** — PR curves, calibration, and metric line plots vs S
5. **By-φ grids** — same structure
6. **By-p_causal grids** (if `model = "sparse_inf"`)

See [`docs/evaluation.md`](../docs/evaluation.md) for the precise formula for
every metric and the structure of every plot.

---

# 9. Where to go next

You now have a working end-to-end pipeline. Common next steps:

- **Larger simulations.** Use `prepare_vcfs.R` to download 50 real 1000G
  regions, then pass `vcf_dir = "data/vcf"` to `run_simulation()`. See
  [README §Reference data](../README.md#reference-data-optional).
- **More methods.** Tier 2 adds FINEMAP and PAINTOR (C++ binaries); Tier 3
  adds the Python-based methods (Funmap, BEATRICE, Functional BEATRICE).
  See [README §Installation](../README.md#installation).
- **Annotation-aware methods.** Re-run with `annotations = "binary"` and
  add `polyfun_oracle`, `polyfun_est`, `funmap`, or `paintor` to the methods
  list. See [`docs/methods.md`](../docs/methods.md) for per-method details.
- **Genome-wide simulation.** For a more realistic experiment (single shared
  phenotype across many regions, polygenic causal assignment), use
  `simulate_gwfm_data()` instead of `run_simulation()`. See
  [`docs/gw_simulation_documentation.md`](../docs/gw_simulation_documentation.md).
- **HPC runs.** For large parameter sweeps, the `scripts/hpc/` directory has
  a SLURM job array template. See
  [README §Running on an HPC cluster](../README.md#running-on-an-hpc-cluster).
- **Evaluation internals.** For the precise definition of every metric and
  the SE-across-replicates machinery, see
  [`docs/evaluation.md`](../docs/evaluation.md).

---

# Session info

```{r session-info}
sessionInfo()
```
